---
title: "TAZ-level Multiplier with regional variables"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'taz_regional_multiplier.html'))})
author: |
  | Simon Brewer
  | Geography Department
  | University of Utah
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float: yes
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(ggdag)
library(lme4)
library(lmerTest)
library(sjPlot)
library(piecewiseSEM)
library(skimr)

load("./data/all_data.RData")
```

## Background

The overall goal is to estimate a multiplier representing the effects of transit on vehicle miles traveled (VMT). The multiplier is comprised of direct (`ADE`) and indirect (`AIE`) effects:

\[
TM = (ADE + AIE) / ADE
\]

## Regional variable

```{r echo=FALSE}
# reg_var = 'fuel_msa10_400'
reg_var = 'tpm'
dat$reg_var = dat[, reg_var]
dat$reg_var = (dat$reg_var - mean(dat$reg_var)) / sd(dat$reg_var)
dat$transittrips <- ifelse(dat$transittrips > 2, 2, dat$transittrips)
# dat$transittrips <- log(dat$transittrips+1)
# dat <- dat[-5093, ]
skim(dat$tpm)
```

## Model paths

Ignoring confounders (to keep things simple) we can develop the following set of paths or structural equation model (we use `transit_acc` to represent accessibility and `act_den` to represent activity density). This is the directed acyclic graph (*DAG*) for the main components

A rough illustration of these given below:

```{r echo=FALSE, fig.height=8}
multi_dag <- dagify(anytransit ~ transit_acc,
                    transittrips ~ transit_acc + anytransit,
                    act_den ~ transit_acc,
                    vmt ~ transit_acc + act_den + transittrips,
                    coords = list(x = c(transit_acc = 1, 
                                        anytransit = 2,
                                        transittrips = 2,
                                        act_den = 2,
                                        vmt = 3),
                                  y = c(transit_acc = 2, 
                                        anytransit = 4,
                                        transittrips = 3,
                                        act_den = 1,
                                        vmt = 2)
                    )
)

multi_dag = multi_dag %>%
  tidy_dagitty() %>%
  mutate(type = c('blue', 'red', 'blue', 'red', 
                  'red', 'blue', 'red', 'red'))

multi_dag2 <- multi_dag

int_df <- data.frame(name = c(NA, NA, NA, NA),
                     x = c(2, 1, 2, 2),
                     y = c(1, 2, 3, 3),
                     direction = c(multi_dag$data$direction[1]),
                     to = c(NA, NA, NA, NA),
                     xend = c(2.5, 2.5, 2.5, 2),
                     yend = c(2.7, 2.7, 1.2, 2),
                     circular = FALSE,
                     type = c('black', 'black', 'black', 'black'))

text_df <- data.frame(x = c(1.6, 2.1, 2.5,
                            1.6, 1.5, 2.5,
                            2, 
                            2.45, 2.25, 2.45, 1.95,
                            2.55, 2.62), 
                      y = c(3.5, 3.5, 2.7,
                            2.75, 1.3, 1.3,
                            2.10, 
                            2.4, 2.7, 1.6, 2.65, 
                            2.85, 1.22),
                      label = c("m1,1", "m2,1", "m4,1",
                                "m2,2", "m3,1", "m4,2",
                                "m4,3",
                                "i1", "i2", "i3", "i4",
                                "r1", "r2")
)

reg_df <- data.frame(x = c(2.5, 2.5, 2.7, 3), 
                     y = c(4, 3, 1, 2.5),
                     label = c("RegVar1", "RegVar2", "RegVar3", "RegVar4")
)

reg_int_df <- data.frame(name = NA,
                     x = c(2.5, 
                           2.5, 2.5,
                           2.7, 2.7,
                           3),
                     y = c(4, 
                           3, 3.05,
                           1, 1,
                           2.5),
                     direction = c(multi_dag$data$direction[1]),
                     to = NA,
                     xend = c(2.05, 
                              2.05, 2.5,
                              2, 2.45,
                              3),
                     yend = c(4, 
                              3, 2.6,
                              1, 1.35,
                              2),
                     circular = FALSE,
                     type = 'darkgreen')

multi_dag2 <- rbind(multi_dag2$data, int_df, reg_int_df)

multi_dag2 %>%
  # tidy_dagitty() %>%
  ggplot() +
  geom_dag_node(aes(x = x, y = y), color = 'white') +
  geom_dag_edges(aes(x = x, y = y, xend = xend, yend = yend, edge_colour = type)) +
  geom_dag_text(aes(x = x, y = y), color = 'black') +
  geom_text(data = text_df, aes(x = x, y = y, label = label)) +
  geom_label(data = reg_df, aes(x = x, y = y, label = label)) +
  theme_dag()
```

This shows that the basic multiplier is based on a set of <span style="color:red">direct</span> and <span style="color:blue">indirect</span> paths: 

- Direct effects
    - `transit accessibility` $\rightarrow \beta_{m1,1} \rightarrow$ `anytransit` $\rightarrow \beta_{m2,1} \rightarrow$ `transittrips` $\rightarrow \beta_{m4,1} \rightarrow$ `vmt`
    - `transit accessibility` $\rightarrow \beta_{m2,2} \rightarrow$ `transittrips` $\rightarrow \beta_{m4,1} \rightarrow$ `vmt`
- Indirect effects
    - `transit accessibility` $\rightarrow \beta_{m3,1} \rightarrow$ `activity density` $\rightarrow \beta_{4,2} \rightarrow$ `vmt`
    - `transit accessibility` $\rightarrow \beta_{m4,3} \rightarrow$ `vmt`

Where $\beta_{k,m}$ is the path coefficient from a regression model. From this we can estimate the two effects for the entire region:

- Direct:

$$
\begin{aligned}
ADE & = (\beta_{m1,1} \times \beta_{m2,1} \times \beta_{m4,1}) + (\beta_{m2,2} \times \beta_{m4,1})
\end{aligned}
$$

- Indirect: 

$$
\begin{aligned}
AIE = (\beta_{m3,1} \times \beta_{m4,2}) + \beta_{m4,3}
\end{aligned}
$$

To allow variations in the multiplier at the TAZ level, we need to include some interactions (or *moderators*) that will adjust coefficients and ultimately the multiplier for a given TAZ. 

- `i1`: the effect of `transittrips` on `vmt` is moderated by activity density
- `i2`: the effect of `transittrips` on `vmt` is moderated by transit accessibility
- `i3`: the effect of activity density on `vmt` is moderated by `transittrips`
- `i4`: the effect of transit accessibility on `vmt` is moderated by `transittrips`
- `r1`: the effect of `transittrips` on `vmt` is moderated by regional population
- `r2`: the effect of activity density on `vmt` is moderated by regional population

And using these, we can update the equations to get TAZ-level direct and indirect effects that include regional variation:

- Direct: 

$$
\begin{split}
ADE = & (\beta_{m1,1} \times \beta_{m2,1} \times (\beta_{m4,1} + i_1 \cdot \mbox{actden} + i_2 \cdot \mbox{pctemp30t} + r_1 \cdot \mbox{reg_var})) \: + \\
& (\beta_{m2,2} \times (\beta_{m4,1} + i_1 \cdot \mbox{actden} + i_2 \cdot \mbox{pctemp30t} + r_1 \cdot \mbox{reg_var}))
\end{split}
$$

- Indirect: 

$$
\begin{split}
AIE = & (\beta_{m3,1} \times (\beta_{m4,2} + i_3 \cdot \mbox{transittrips}) + r_2 \cdot \mbox{reg_var}) \: + \\
& (\beta_{m4,3} + i_4 \cdot \mbox{transittrips})
\end{split}
$$

## Coefficients

```{r echo=FALSE}
## ------------------------
## Model 1
f1 <- anytransit ~ pctemp30t + hhsize + employed + 
  incn + entropy + pct4way + reg_var +
  (1 | region)
# f1 <- anytransit ~ pctemp30t + hhsize + employed + 
#   pct4way + 
#   (1 | region)
lm1 <- lmer(f1, 
            data = dat,
            REML = TRUE,
            control = lmerControl(optimizer="Nelder_Mead"))

## ------------------------
## Model 2
f2 <- transittrips ~ pctemp30t + hhsize + employed + 
  incn + entropy + anytransit + reg_var +
  (1 | region)
# f2 <- transittrips ~ pctemp30t + hhsize + employed + 
#   incn + entropy + anytransit + 
#   (1 | region)
lm2 <- lmer(f2, 
            data = dat,
            REML = TRUE,
            control = lmerControl(optimizer="Nelder_Mead"))

## ------------------------
## Model 3
f3 <- actden ~ pctemp30t + entropy + pct4way +
  pctemp30a + reg_var + 
  (1 | region)
# f3 <- actden ~ pctemp30t + entropy + pct4way +
#   pctemp30a + 
#   (1 | region)
lm3 <- lmer(f3, 
            data = dat,
            REML = TRUE,
            control = lmerControl(optimizer="Nelder_Mead"))

## ------------------------
## Model 4
f4 <- vmt100occ ~ hhsize + employed + incn +
  entropy + pct4way +
  transittrips * actden + transittrips * pctemp30t + 
  reg_var * actden + reg_var * transittrips +
  (1 | region)
lm4 <- lmer(f4,
            data = dat,
            REML = TRUE,
            control = lmerControl(optimizer="Nelder_Mead"))
```

### Path coefficients

```{r}
m11 = as.numeric(fixef(lm1)['pctemp30t'])
m21 = as.numeric(fixef(lm2)['anytransit'])
m22 = as.numeric(fixef(lm2)['pctemp30t'])
m41 = as.numeric(fixef(lm4)['transittrips'])
```

```{r}
m31 = as.numeric(fixef(lm3)['pctemp30t'])
m42 = as.numeric(fixef(lm4)['actden'])
m43 = as.numeric(fixef(lm4)['pctemp30t'])
```

```{r}
coef_df = data.frame(m11, m21, m22, m31, m41, m42, m43)
knitr::kable(coef_df)
```

### Interactions

```{r}
i1 <- as.numeric(fixef(lm4)['transittrips:actden'])
i2 <- as.numeric(fixef(lm4)['transittrips:pctemp30t'])
i3 <- as.numeric(fixef(lm4)['transittrips:actden'])
i4 <- as.numeric(fixef(lm4)['transittrips:pctemp30t'])
r1 <- as.numeric(fixef(lm4)['transittrips:reg_var'])
r2 <- as.numeric(fixef(lm4)['actden:reg_var'])
```

```{r}
int_df = data.frame(i1, i2, i3, i4, r1, r2)
knitr::kable(int_df)
```

### Final direct effect equation

$$
\begin{split}
ADE & = ({`r round(m11, 4)`} \times {`r round(m21, 4)`} \times ({`r round(m41, 4)`} + {`r round(i1, 4)`} \cdot \mbox{actden} + {`r round(i2, 4)`} \cdot \mbox{pctemp30t} + {`r round(r1, 4)`} \cdot \mbox{reg_var})) \\
& + ({`r round(m22, 4)`} \times ({`r round(m41, 4)`} + {`r round(i1, 4)`} \cdot \mbox{actden} + {`r round(i2, 4)`} \cdot \mbox{pctemp30t} + {`r round(r1, 4)`} \cdot \mbox{reg_var}))
\end{split}
$$

### Final indirect effect equation

$$
\begin{aligned}
AIE & = ({`r round(m31, 4)`} \times ({`r round(m42, 4)`} + {`r round(i3, 4)`} \cdot \mbox{transittrips} + {`r round(r2, 4)`} \cdot \mbox{reg_var})) \\
& + ({`r round(m43, 4)`} + {`r round(i4, 4)`} \cdot \mbox{transittrips})
\end{aligned}
$$

### TAZ Multipliers

- Direct effect

```{r}
ade = as.numeric((m11 * m21 * (m41 + i1 * dat$actden + i2 * dat$pctemp30t + r1 * dat$reg_var)) + 
                   (m22 * (m41 + i1 * dat$actden + i2 * dat$pctemp30t + r1 * dat$reg_var)))
print(median(ade))
```

- Indirect effect

```{r}
aie = as.numeric((m31 * (m42 + i3 * dat$transittrips + r2 * dat$reg_var)) + (m43 + i4 * dat$transittrips))
print(median(aie))
```

- Multiplier

```{r}
TM = (ade + aie) / ade
print(range(TM))
```

```{r}
#TM2 <- TM[TM < quantile(TM, 0.99)]
# TM2 <- TM[which(TM > 1)]
hist(TM, breaks = 100)
```

## Placetype Multiplier

```{r}
dat$TM = TM
dat_pt = dat %>%
  filter(PlaceType != "0")
```

```{r}
ggplot(dat_pt, aes(x = PlaceType, y = TM)) +
         geom_boxplot() +
  scale_y_log10() +
  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  theme_bw()
```

```{r}
ggplot(dat_pt, aes(x = reorder(region, TM, median), y = TM)) +
         geom_boxplot() +
  scale_y_log10() +
  scale_x_discrete("Region", guide = guide_axis(n.dodge = 2)) +
  theme_bw()
```


```{r message=FALSE, warning=FALSE}
dat_reg = dat_pt %>%
  group_by(PlaceType, region) %>%
  summarize(actden = mean(actden),
            pctemp30t = mean(pctemp30t),
            transittrips = mean(transittrips),
            reg_var = mean(reg_var))

ade = as.numeric((m11 * m21 * 
                    (m41 + i1 * dat_reg$actden + 
                       i2 * dat_reg$pctemp30t + r1 * dat_reg$reg_var)) + 
                   (m22 * (m41 + i1 * dat_reg$actden + 
                             i2 * dat_reg$pctemp30t + 
                             r1 * dat_reg$reg_var)))

aie = as.numeric((m31 * (m42 + i3 * dat_reg$transittrips + 
                           r2 * dat_reg$reg_var)) + 
                   (m43 + i4 * dat_reg$transittrips))

dat_reg$TM = (ade + aie) / ade
```

```{r message=FALSE, warning=FALSE}
library(reshape2)
dat_reg_tab = dcast(dat_reg, region ~ PlaceType, value.var = "TM")
dat_reg_tab = dat_reg_tab %>% 
  ungroup() %>% 
  # mutate(Total = rowSums(across(where(is.numeric)))) %>%
  bind_rows(summarize(., region = "Mean", 
                      across(where(is.numeric), \(x) mean(x, na.rm = TRUE))))

knitr::kable(dat_reg_tab)
```

```{r}
dat_out <- dat %>%
  select(TAZ, region, region_id, regTAZ, 
         pctemp30t, actden, transittrips, 
         hhsize, incn, employed, entropy, 
         pct4way, PlaceType, reg_var, 
         taz_pop, taz_emp,
         TM)
write.csv(dat_out, "./output/multiplier_output.csv", row.names = FALSE)
```

```{r}
dat_out <- dat %>%
  select(TAZ, region, region_id, regTAZ, 
         PlaceType, 
         TM)
write.csv(dat_out, "./output/taz_multiplier_output.csv", row.names = FALSE)
```

### Regional Multiplier

- Direct effect

```{r}
ade = as.numeric((m11 * m21 * m41) + (m22 * m41))
print(median(ade))
```

- Indirect effect

```{r}
aie = as.numeric((m31 * m42) + m43)
print(median(aie))
```

- Multiplier

```{r}
TM = (ade + aie) / ade
print(TM)
```

## Appendix 1: Models


### Model 1 `anytransit`

```{r echo=FALSE}
tab_model(lm1, digits = 4)
```

### Model 2: `transittrips`

```{r echo=FALSE}
tab_model(lm2, digits = 4)
```

### Model 3: `actden`

```{r echo=FALSE}
tab_model(lm3, digits = 4)
```

### Model 4: `vmt100occ`

```{r echo=FALSE}
tab_model(lm4, digits = 4)
```


